Spatial distribution of environmental DNA in a nearshore marine habitat

### 2017-01-07 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex

\bigskip
% }

We are grateful to the editor and reviewers for their insightful feedback.
Reviewer feedback is below in regular typeface, with our responses indented and colored.
% Additional changes to the manuscript independent of the reviews are presented at the end of this document.

%------------------------------------EDITOR-------------------------------------
\subsection*{Editor}
Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to. I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer 1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence). Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative. I wondered what the community analyses results would look like if you compartmentalised by e.g. life history or type (e.g. fish)?

\begin{response}
  Thank you for the positive feedback.
	Indeed, we agree with the comment made on the preprint.
	Our omission was not intended as a slight, merely an oversight given the rapid pace of publication in this field.
	We have serious qualms about the quality of the data in their paper that we believe preclude us from recommending it as a study of spatial variation in eDNA.
	However, the suggestions of reviewer 1 (Wangensteen) were justified and we have added the appropriate citations and corrected our wording.
	% TODO:  respond to comment, update context in paper.
	\\

	Like you, we were curious about the effect of the data subset you mention, and conducted that analysis using the life history data. Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text.
	% TODO: comment on effect of subsetting based on life-history strategy or taxon
	\\
\end{response}

@@

L285. ``333,537.9 ± 112,200.5''. It looks somehow awkward to use decimal values for numbers of reads here. You could round the values to ``333,538 ± 112,200''
\begin{response}
  Because these are summary statistics (mean and standard deviation), we believe the use of decimals is appropriate.\\
\end{response}

L314-315. You can delete ``the probability that two reads chosen at random from a sample belong to different species'', since it has been already explained in the Materials and Methods section.
\begin{response}
  We made this change.\\
\end{response}

L344. ``For the first time in a marine environment, we document four key patterns''. The authors should definitely use a more modest sentence here. At least the first two patterns they claim to report ``for the first time'' have been already reported previously in several works in the marine environment. E.g.: Fonseca et al. 2014, Guardiola et al. 2015, Lallias et a. 2015, De Vargas et al. 2015, etc., although possibly never with a sampling design specifically designed to get a quantitative assessment of distance decay.
\begin{response}
  Our intent was precisely to convey the novelty of the sampling design; we failed to do that clearly.
	We removed `For the first time in a marine environment', and we have clarified the novelty of our study in the context of recently published work in the introduction.\\
	% TODO: Comment on the novelty of this study in the context of refs: Fonseca et al. 2014, Guardiola et al. 2015 (PLOSONE), Lallias et a. 2015, De Vargas et al. 2015, Hanfling 2016, Guardiola 2016 PeerJ
\end{response}

@@

I felt that under the remit of the Ecological Analyses headings throughout the study, there could have been some focus on reporting the alpha diversity and rarefaction analysis in relation to whether effective sequencing depth had been achieved for the samples in question.
\begin{response}
	% TODO: report alpha diversity and rarefaction analysis to demonstrate effective sequence depth had been achieved
  OurResponse\\
\end{response}

\subsubsection*{Comments for the Author}
I thought the paper was very interesting. The study was well designed, implemented, and written. It was useful to clarify the expectations and hypotheses outlined within, however, I felt more effort could have been made to provide recommendations on how some of the limitations in eDNA sampling could be overcome.
\begin{response}
	% TODO: provide recommendations on how some of the limitations in eDNA sampling could be overcome
  OurResponse\\
\end{response}

The bioinformatics/statistical methods seemed to be very well thought out, and consistent with current best practices in the analyses of this type of sequencing data.
\begin{response}
	Thank you; we agree.\\
\end{response}

%------------------------------------REVIEWER3----------------------------------
\subsection*{Comment on the preprint}

\subsubsection*{Bernd Hänfling, Lori Lawson Handley, Dan Read and Ian Winfield}

This is well designed study and we would like to congratulate the authors to this data set and its excellent analysis. However, we feel that the authors have made an important omission in the discussion.\\

@@

Given that the main focus of this paper is to explore the spatial distribution of community eDNA we feel that the authors should consider the results of our study on lake fish communities (Hänfling et al. 2016). In our paper we have used extensive spatial sampling of three transects across our main study lake with a total of 66 individual samples and to our knowledge there is no other data set on macrobial eDNA metabarcoding with a comparable extent of explicit sampling. Our paper confirms that the spatial distribution of individual species reflects their abundance but also shows that the pattern of distribution is not random and ecologically meaningful. Although there might be important differences between freshwater and marine habitats we feel that the results from our study are extremely relevant here and should not be omitted especially when claiming novelty of discussion point 2.\\

Hänfling B, Lawson Handley L, Read DS, Hahn C, Li J, Nichols P, Blackman RC, Oliver A, Winfield IJ (2016) Environmental DNA metabarcoding of lake fish communities reflects long-term data from established survey methods. Molecular Ecology 25, 3101-3119.
\begin{response}
	% http://onlinelibrary.wiley.com/doi/10.1111/mec.13660/full
  Thank you for your comment and for drawing our attention to your paper.
	As you note therein, the rate of publication in this field can make it difficult to keep up.
	Our omission was not intended as a slight, merely an oversight. % given the rapid pace of publication in this field.
	% We acknowledge that your study took place in a spatial context becuase you analyzed samples collected from transects in a lake.
	We commend your extensive environmental sampling; however, upon reading your paper, we have reservations about your interpretations of the results given the molecular protocols. % are reluctant to
  \\

	We note that you conducted replicate PCRs of each environmental sample, and pooled them prior to sequencing.
	While this practice is not uncommon in the field, it precludes the ability to test for any effects of PCR on the communities.
	We believe this is particularly problematic in the case of your study because, as far as we understand your methods, you used indexed ('tagged') primers in a single PCR reaction.
	We have shown previously that this can bias datasets in such a way that completely obscures any underlying pattern (see: http://dx.doi.org/10.1371/journal.pone.0148698).
	As we report in that study, indexed primers can bias relative read abundance by as much as 70\%.
	That is, in one environmental sample, a species may comprise 70\% of the reads in PCR replicates amplified with one indexed primer set, while being completely absent from replicates amplified by another indexed primer set.
	This would most certainly result in a significant difference among most any comparison, such as the Chi-squared test you conducted to compare lake basins.
	If we have misunderstood your methodology, or if you have results of replication tests on the indexed primers in community samples, we would be happy to consider further the results of your study.
  \\

	Regardless, we have neutralized our statements on the novelty of our work on the basis of points brought up by one of the reviewers.

	% In your paper, you note that `the results regarding vertical sampling should be regarded as preliminary' and `Our site occupancy estimates should therefore be treated as preliminary'
	% Yet these site occupancy estimates are the data on which the
	% Note that even for the mock community, we found no formal analysis of whether read count or relative abundance was correlated with input DNA mass.
	% Sampling design complicates the relationship between distance and community similarity because depth and distance are confounded.
\end{response}

### 2017-01-09 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex

\documentclass{article}
\usepackage{changepage}% \adjustwidth{<leftskip>}{<rightskip>} http://ctan.org/pkg/changepage
\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
\usepackage[utf8x]{inputenc} % Allow input of accented characters (and more...)
\usepackage[usenames, dvipsnames]{xcolor} % to use color
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides

@@

\bigskip
% }

We are grateful to the editor and reviewers for their insightful feedback.
Reviewer feedback is below in regular typeface, with our responses indented and colored.
% Additional changes to the manuscript independent of the reviews are presented at the end of this document.

%------------------------------------EDITOR-------------------------------------
\subsection*{Editor}
Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to. I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer 1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence). Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative. I wondered what the community analyses results would look like if you compartmentalised by e.g. life history or type (e.g. fish)?

\begin{response}
  Thank you for the positive feedback.
	Indeed, we agree with the comment made on the preprint that we overlooked some recently published works.
	We have serious qualms about the quality of the data in their paper that we believe preclude us from recommending it as a study of spatial variation in eDNA.
	However, the suggestions of reviewer 1 (Wangensteen) were justified and we have added the appropriate citations and corrected our wording.
	% TODO:  respond to comment, update context in paper.
	\\

### 2017-01-09 Manuscript/v1.1_PeerJ/carkeek_grid.tex
respond to reviews
more responding to reviews

While rigorous laboratory studies have investigated the effects of some environmental factors on eDNA persistence \citep{Klymus2015, Barnes2014, Sassoubre2016} and the transport of eDNA in specific contexts \citep{Deiner2014}, we suggest that field studies comparing the spatial distribution of communities of eDNA with expectations based on prior knowledge of organisms' distributions are also critical to developing a working understanding of eDNA in the real world.

@@

The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases. Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}. Both have undesirable effects of biasing estimates of diversity: in our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample. In turn, this could obscure true underlying patterns of diversity. In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result. In fact, even surveys relying on traditional capture techniques (e.g. seine nets) and morphological characteristics are subject to biases imposed by the sampling gear (e.g. mesh size), the observer (e.g. taxonomic expertise), and organisms (e.g. morphologically cryptic species). Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods \citep{Kelly2016FMS}.

>>>>

While rigorous laboratory studies have investigated the effects of some environmental factors on eDNA persistence \citep{Klymus2015, Barnes2014, Sassoubre2016} and the transport of eDNA in specific contexts \citep{Deiner2014}, we suggest that field studies comparing the spatial distribution of communities of eDNA with expectations based on prior knowledge of organisms' distributions are also critical to developing a working understanding of eDNA in the real world.
Research to date has documented the non-random spatial distribution of % spatial variation of
meiofaunal \citep{Fonseca2014,Guardiola2016}, microbial \citep{Lallias2015}, and extracellular \citep{Guardiola2015} eDNA of marine and estuarine sediments, and of microscopic plankton in open ocean waters \citep{DeVargas2015}.
These studies conducted targeted sampling at intermediate (tens of kilometers) to global (thousands of kilometers) scales.
Here, we use a grid-based environmental sampling strategy to assess spatial variability of eDNA in a coastal marine environment at a fine scale (tens to thousands of meters), using molecular methods that focus on macrobial metazoans.

@@

The marker we chose for this study detects a wide variety of metazoans while excluding other more common taxa; however, it does not effectively discriminate among species within a higher group in all cases.
Other markers, such as mitochondrial cytochrome c oxidase subunit 1 (COX1, CO1, or COI) may provide adequate species-level resolution in some metazoan groups, but have other shortcomings including taxon dropout \citep{Deagle2014} and amplification of more abundant non-metazoans, as we discovered in an accompanying study \citep{Kelly2016FMS}.
Both have undesirable effects of biasing estimates of diversity.
In our case, it is possible that the lumping of multiple species into one group underestimates the true richness of the group and of the entire sample, in turn obscuring true underlying patterns of diversity.
In the case of COX1, well-documented primer biases cause failure to amplify some taxa, particularly in mixed samples, with the same result \citep{Deagle2014}.
In fact, even surveys relying on traditional capture techniques (e.g. seine nets) and morphological characteristics are subject to biases imposed by the sampling gear (e.g. mesh size), the observer (e.g. taxonomic expertise), and organisms (e.g. morphologically cryptic species).
Similarly, no single molecular marker adequately and effectively samples all taxa without bias  \citep{Drummond2015}, and thus the choice of marker is an important and context-dependent one.
Until whole-genome sequencing of individual cells is a reality, the tradeoffs between taxonomic breadth and resolution will continue to be problematic for metabarcoding studies, just as they are for more traditional ecological survey methods \citep{Kelly2016FMS}.
